---
title: "Dealing with Dirty Data Final Project"
author: "Michael Wakeman"
date: "1/9/2021"
output: html_document
---

```{r setup, include=FALSE}
#Initial packages and loading of original data file using readr.

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(naniar)

Original.Data <- read_csv("NYC_Wi-Fi_Hotspot_Locations.csv")
Wifi <- Original.Data

```

## Introduction

For my final project, I chose to utilize the "NYC Wi-Fi Hotspot Locations" data set from [NYC Open Data](https://data.cityofnewyork.us/City-Government/NYC-Wi-Fi-Hotspot-Locations/yjub-udmw), which was retrieved on Jan. 7, 2021 as a .csv file. The data set includes information on hospot wifi within NYC, including their location, type, and provider information. I chose to call this data set `Wifi`.

I would like to learn about the geographic distribution of hotspot locations in NYC, including by neighborhood and burough. This would also include where internet hotspots are most available and least available. I would also like to know about accessibility to these hotspots, are they mostly free?

## Data Exploration

Here are the initial variables in the `Wifi` data set. I chose to rename `Council Distrcit` to `Council District` since it appeared to be a spelling mistake. There are 29 columns and 3319 rows in the data set. 

```{r Variable Exploration}
#Code used to find variable names and the amount of rows and columns.

colnames(Wifi)
ncol(Wifi)
nrow(Wifi)

```

```{r, include = FALSE}
#Renaming variable due to spelling error

colnames(Wifi)[21] <- "Council District"

```

### Variable Definitions

Here is a table documenting the original variables and their definitions taken from NYC Open Data.

```{r Documentation, echo = FALSE}
#Creating Data Table of variables and renaming columns

Document.variables <- data.frame(colnames(Wifi), "Definition")
colnames(Document.variables)[1] <- "Variable"
colnames(Document.variables)[2] <- "Definition"

#Recoding definitions per NYC Open Data

Document.variables[1, 2] <- "Identification number automatically generated by map software ArcMap."
Document.variables[2, 2] <- "Borough of New York City. MN = Manhattan BX = Bronx BK = Brooklyn QU = Queens SI = Staten Island."
Document.variables[3, 2] <- "Type of WiFi provided by franchise."
Document.variables[4, 2] <- "Franchise who is providing the Wifi connection."
Document.variables[5, 2] <- "The name of the location where the WiFi is located."
Document.variables[6, 2] <- "A brief description of where the WiFi point is."
Document.variables[7, 2] <- "Points that fall North or South of the Equator, expressed in degrees."
Document.variables[8, 2] <- "Points that fall East or West of the Prime Meridian, expressed in degrees."
Document.variables[9, 2] <- "A horizontal value where a point is located."
Document.variables[10, 2] <- "A vertical value where a point is located."
Document.variables[11, 2] <- "The type of location that a WiFi hotspot is present in."
Document.variables[12, 2] <- "Comments describing a specific hotspot or location."
Document.variables[13, 2] <- "The city in which a hotspot is located."
Document.variables[14, 2] <- "The name of the WiFi seen on people's devices."
Document.variables[15, 2] <- "A LinkNYC Kiosk's unique identification number."
Document.variables[16, 2] <- "The approximated date that this Wi-Fi site went active."
Document.variables[17, 2] <- "The New York City borough where the hotspots are located."
Document.variables[18, 2] <- "The New York City borough where the hotspots are located."
Document.variables[19, 2] <- "Four (4) digit code assigned to neighborhoods to differentiate NYC's five boroughs."
Document.variables[20, 2] <- "Neighborhood identifiers to differentiate NYC's five boroughs."
Document.variables[21, 2] <- "NYC City Council District number."
Document.variables[22, 2] <- "United States Postal Service Zip Code (5-digits)."
Document.variables[23, 2] <- "New York City Borough plus Community Board Numbers."
Document.variables[24, 2] <- "2010 U.S. Census data based on location."
Document.variables[25, 2] <- "Combined value of Borough CD and Census 2010."
Document.variables[26, 2] <- "Building Identification Number."
Document.variables[27, 2] <- "Borough Block Lot."
Document.variables[28, 2] <- "Identification number automatically generated by GIS Developer map sites."
Document.variables[29, 2] <- "Combination of the lat and long values."

#Final paged output

rmarkdown::paged_table(Document.variables)

```

## Initial Data Recoding

#set 0 to NA for sourceid, name, bin, bbl
#activated has extremely weird dates still

The data includes `NA` values within several variables. In addition, `BBL`, `BIN`, `Name`, and `SourceID` contain observations of 0. Given their definitions these are most likely missing values as well. For that reason I recoded `0` within each of these columns as `NA`.

Additionally, `Activated` contains dates. This includes values of `09/09/9999` which I also recoded as `NA`.

```{r Missing Data Recoding, echo = FALSE}
#Recoding values as NA for several variables

Wifi <- Wifi %>%
  naniar::replace_with_na(replace = list(BBL = 0,
                                         Activated = "09/09/9999",
                                         BIN = 0,
                                         Name = 0,
                                         SourceID = 0))
```


## Initial Data Cleaning


```{r Missing Data Binary Flags, echo = FALSE}
#Binary flags for every variable based on missing data using ifelse statement. A missing (NA) observation will appear as a 1.

Wifi$BF_OBJECTID <- ifelse(is.na(Wifi$OBJECTID) == TRUE, 1, 0)
Wifi$BF_Provider <- ifelse(is.na(Wifi$Provider) == TRUE, 1, 0)
Wifi$BF_Latitude <- ifelse(is.na(Wifi$Latitude) == TRUE, 1, 0)
Wifi$BF_Y <- ifelse(is.na(Wifi$Y) == TRUE, 1, 0)
Wifi$BF_City <- ifelse(is.na(Wifi$City) == TRUE, 1, 0)
Wifi$BF_Activated <- ifelse(is.na(Wifi$Activated) == TRUE, 1, 0)
Wifi$"BF_Neighborhood Tabulation Area Code (NTACODE)" <- 
  ifelse(is.na(Wifi$"Neighborhood Tabulation Area Code (NTACODE)") == TRUE, 1, 0)
Wifi$BF_Postcode <- ifelse(is.na(Wifi$Postcode) == TRUE, 1, 0)
Wifi$BF_BCTCB2010 <- ifelse(is.na(Wifi$BCTCB2010) == TRUE, 1, 0)
Wifi$BF_DOITT_ID <- ifelse(is.na(Wifi$DOITT_ID) == TRUE, 1, 0)
Wifi$BF_Borough <- ifelse(is.na(Wifi$Borough) == TRUE, 1, 0)
Wifi$BF_Name <- ifelse(is.na(Wifi$Name) == TRUE, 1, 0)
Wifi$BF_Longitude <- ifelse(is.na(Wifi$Longitude) == TRUE, 1, 0)
Wifi$BF_Location_T <- ifelse(is.na(Wifi$Location_T) == TRUE, 1, 0)
Wifi$BF_SSID <- ifelse(is.na(Wifi$SSID) == TRUE, 1, 0)
Wifi$BF_BoroCode <- ifelse(is.na(Wifi$BoroCode) == TRUE, 1, 0)
Wifi$"BF_Neighborhood Tabulation Area (NTA)" <- 
  ifelse(is.na(Wifi$"Neighborhood Tabulation Area (NTA)") == TRUE, 1, 0)
Wifi$BF_BoroCD <- ifelse(is.na(Wifi$BoroCD) == TRUE, 1, 0)
Wifi$BF_BIN <- ifelse(is.na(Wifi$BIN) == TRUE, 1, 0)
Wifi$"BF_Location (Lat, Long)" <- ifelse(is.na(Wifi$"Location (Lat, Long)") == TRUE, 1, 0)
Wifi$BF_Type <- ifelse(is.na(Wifi$Type) == TRUE, 1, 0)
Wifi$BF_Location <- ifelse(is.na(Wifi$Location) == TRUE, 1, 0)
Wifi$BF_X <- ifelse(is.na(Wifi$X) == TRUE, 1, 0)
Wifi$BF_Remarks <- ifelse(is.na(Wifi$Remarks) == TRUE, 1, 0)
Wifi$BF_SourceID <- ifelse(is.na(Wifi$SourceID) == TRUE, 1, 0)
Wifi$"BF_Borough Name" <- ifelse(is.na(Wifi$"Borough Name") == TRUE, 1, 0)
Wifi$"BF_Council District" <- ifelse(is.na(Wifi$"Council District") == TRUE, 1, 0)
Wifi$"BF_Census Tract" <- ifelse(is.na(Wifi$"Census Tract") == TRUE, 1, 0)
Wifi$BF_BBL <- ifelse(is.na(Wifi$BBL) == TRUE, 1, 0)

```

I created binary flags for missing data within each variable using ifelse statements. Of the 29 variables, `Activated`, `BBL`, `BIN`, `Name`, `Remarks`, and `SourceID` contain missing data. These variables all have several hundred missing observations, so it would not be suitable to drop them. I then removed the unnecessary flags. 

```{r, echo = FALSE, include = FALSE}
#Identifying specific variables with missing data using sums. If the sum is greater than 0 than that row has missing data.

sum(Wifi$BF_OBJECTID)
sum(Wifi$BF_Provider)
sum(Wifi$BF_Latitude)
sum(Wifi$BF_Y)
sum(Wifi$BF_City)
sum(Wifi$BF_Activated)
sum(Wifi$"BF_Neighborhood Tabulation Area Code (NTACODE)")
sum(Wifi$BF_Postcode) 
sum(Wifi$BF_BCTCB2010)
sum(Wifi$BF_DOITT_ID)
sum(Wifi$BF_Borough)
sum(Wifi$BF_Name)
sum(Wifi$BF_Longitude)
sum(Wifi$BF_Location_T) 
sum(Wifi$BF_SSID)
sum(Wifi$BF_BoroCode) 
sum(Wifi$"BF_Neighborhood Tabulation Area (NTA)") 
sum(Wifi$BF_BoroCD) 
sum(Wifi$BF_BIN)
sum(Wifi$"BF_Location (Lat, Long)")
sum(Wifi$BF_Type) 
sum(Wifi$BF_Location) 
sum(Wifi$BF_Remarks)
sum(Wifi$BF_X) 
sum(Wifi$BF_SourceID) 
sum(Wifi$"BF_Borough Name") 
sum(Wifi$"BF_Council District") 
sum(Wifi$"BF_Census Tract") 
sum(Wifi$BF_BBL)
    

#Removing binary flags for variables with no missing data

Wifi <- Wifi %>%
  select(-BF_OBJECTID, 
         -BF_Provider,
         -BF_Latitude,
         -BF_Y,
         -BF_City,
         -"BF_Neighborhood Tabulation Area Code (NTACODE)",
         -BF_Postcode,
         -BF_BCTCB2010,
         -BF_DOITT_ID,
         -BF_Borough,
         -BF_Longitude,
         -BF_Location_T,
         -BF_SSID,
         -BF_BoroCode,
         -"BF_Neighborhood Tabulation Area (NTA)",
         -BF_BoroCD,
         -"BF_Location (Lat, Long)",
         -BF_Type,
         -BF_Location,
         -BF_X,
         -"BF_Borough Name",
         -"BF_Council District",
         -"BF_Census Tract",
         )   

```

I then starting modifying individual variables. I began by recoding `OBJECTID` to begin at `1` instead of `9601`. 

`Borough` and `BoroCode` also contained the same information, so I removed `BoroCode` from the data set. I verified this programatically. 

As mentioned previously, the only date within the data set is `Activated`. I used the Lubridate package to recode this variable as a properly formatted date. 

Finally, `Census Tract` and `BCTCB2010` contained the same information, so I removed the latter column.


```{r Initial Data Cleaning}
#Recoded OBJECTID to begin at 1 instead of 9601 using subtraction.

Wifi <- Wifi %>%
  arrange(OBJECTID) %>%
  mutate(OBJECTID = (OBJECTID - 9600))

#check to see if Borocode and Borough are the same variable. Since the sum returned as 0, they are equal.

bouroughcodeyes <- ifelse(Wifi$Borough == Wifi$BoroCode, 0, 1)
sum(bouroughcodeyes)

#Removed Borocode due to being same data as Borough
Wifi <- Wifi %>%
  select(-BoroCode)

#FOrmatting Activated dates properly using Lubridate

Wifi$Activated <- mdy(Wifi$Activated)

#Also checking if BCT is the same as census tract. This is the same previous check. Since it is the same (sum = 0) I chose to remove it.

censustractcheck <- ifelse(Wifi$`Census Tract` == Wifi$BCTCB2010, 0, 1)
sum(censustractcheck)

Wifi <- Wifi %>%
  select(-BCTCB2010)

```

I then started modifying the `City` and `Borough Name` variables. These two share observations for most rows. However, there are some discrepancies. I analyzed each borough to find these differences. Of the five boroughs, there were no errors with the Bronx, Brooklyn, or Staten Island.

Manhattan is the only borough to correspond with `New York` as the city. The other boroughs have themselves listed as the city. To maintain this pattern I recoded `New York` to `Manhattan` (and NYC is the city for all hotspots anyways). Additionally, there was one occurrence where Manhattan corresponded with `Brooklyn` as a city. I interpreted this as an error and recoded it to `Manhattan` as well. 

Finally, Queens has many interesting corresponding values in `City`. This includes errors, such as other boroughs like `Brooklyn`. It also includes smaller cities like `Long Island City`. These do not always match the corresponding NTA. Additionally, these are only a small subset of the `City` data and only for Queens, so I believe these values are errors. I recoded these values to `Queens`.

Since all of the values for `City` matched `Borough` after recoding, I chose to remove the latter from the data set.

```{r Boroughs, echo = FALSE}
#No mismatches in the Brooklyn borough

brooklyn <- Wifi %>%
  filter(City != `Borough Name`) %>%
  select(City, `Borough Name`, `Neighborhood Tabulation Area Code (NTACODE)`) %>%
  filter(`Borough Name` == "Brooklyn")
brooklyn

#No mismatches for Staten island borough

Staten_Island <- Wifi %>%
  filter(City != `Borough Name`) %>%
  select(City, `Borough Name`, `Neighborhood Tabulation Area Code (NTACODE)`) %>%
  filter(`Borough Name` == "Staten Island")
Staten_Island

#No mismatches in Bronx borough

Bronx <- Wifi %>%
  filter(City != `Borough Name`) %>%
  select(City, `Borough Name`, `Neighborhood Tabulation Area Code (NTACODE)`) %>%
  filter(`Borough Name` == "Bronx")
Bronx

#There is one where Brooklyn is encoded to Manhattan. Otherwise New York in city is encoded to Manhattan borough.

Manhattan <- Wifi %>%
  filter(City != `Borough Name`) %>%
  select(City, `Borough Name`, `Neighborhood Tabulation Area (NTA)`, OBJECTID) %>%
  filter(`Borough Name` == "Manhattan") %>%
  filter(City != "New York")
Manhattan

#Recoding values that don't match Manhattan borough

Wifi$City[Wifi$City == "New York"] <- "Manhattan"
Wifi[1263, 13] <- "Manhattan"

#QUeens has many incorrect values. These values for City are similar to NTA but slightly different. This includes setting Brooklyn or other boroughs within the Queens borough.

queens <- Wifi %>%
  filter(City != `Borough Name`) %>%
  select(City, `Borough Name`, `Neighborhood Tabulation Area (NTA)`, `Neighborhood Tabulation Area Code (NTACODE)`) %>%
  filter(`Borough Name` == "Queens")
queens

#Recoding city areas from Queens


```

 Lastly, I recoded several categorical variables as factors. These variables were `Borough`, `City`, `Provider`, and `Type`.
 
```{r}
#Recoding categorical variables as factors

Wifi$Borough <- as.factor(Wifi$Borough)
Wifi$City <- as.factor(Wifi$City)
Wifi$Type <- as.factor(Wifi$Type)
Wifi$Provider <- as.factor(Wifi$Provider)

```

### String Cleaning

```{r, eval = FALSE}
#Now Im gonna split up the location into two new variables. It seems to include street number and street address, so basically just what we did with the exercise.

#first im going to make an address variable

#attempt$`Mailing Address` <- str_c(attempt$Location, attempt$Cit, "New York", attempt$Postcode, sep = ", ")

#Now I can separate out the street number and street address
#I have to filter out the ones with actual addresses with numbers (instead of getting "Skate Rental Area' for example)
##address <- str_split(attempt$Location, pattern = " ", n = 2, simplify = TRUE)

#Yeah theres a lot of differences in this variable I'll have to sort through still
#SO ill have to modify mailing address to be for the ones with numbers in the address
#It looks like some of these contain whole addresses too with zip and state too so I'll have to remove that as well


```

```{r pressure, echo=FALSE}
#Break apart remarks to get the speed/quality


```

## Variable Formation

```{r}
#Binary variable flag for free Wifi

Wifi$Free <- ifelse(Wifi$Type == "Free", 1, 0)


unique(Wifi$Location_T)
#Create binary indoor outdoor variable

```

## Analysis

```{r}
#Want to see what the most common provider is by borough, NTA
#Want to see what the most common type is by borough, NTA
#Want to see what borough, NTA has the most free hotspots
```

## Plots

```{r}
#Bar graph for type per NTA, borough

```

